Safe Transferable Regions

### 2015-11-17 paper/overview2.tex
Edits to Sec.2.

\paragraph{Stack Regions}
The ``\C{letregion R \{ S \}}'' construct creates a new stack region,
with a static identifier \C{R}, whose scope is restricted to the
statement \C{S}. The semantics of  \C{letregion} is similar
to Tofte and Talpin~\cite{tofte94}'s \C{letregion} expression:
new objects can be allocated in the newly created region within
the scope of \C{S}, but the region and all objects allocated within it
are freed at the end of \C{S}.

@@

The \C{Region} constructor takes as a parameter a function that constructs
the root object: it creates a new region and invokes this function, with the new
region as the allocation context, to create the root object of the region.
This following code illustrates the creation of a transferable region:
% (\C{r}) of type \C{T}:
\begin{codejava}
  Region<T> r = new Region<T>(() => new T())
\end{codejava}

\begin{figure}
\begin{codejava}
  public class Region<T> {
    public Region(Func<T> mkRoot);
    public void free();
    public void transfer();
  }
\end{codejava}
\caption{The type signature of the \C{Region} class}
\label{fig:region-class}
\end{figure}

The type signature of the class \C{Region} appears in Fig.~\ref{fig:region-class}.
The method \C{free} deallocates the region (including all objects allocated within it).
The method \C{transfer} transfers the region to a downstream actor as determined by the run-time.
It is an abstraction of two possible forms of  transfer: a transfer between two actors in a shared
memory setting or a transfer between two actors in a distributed, message-passing, setting. 
The precise semantics of \C{transfer} are unimportant in the context of the region type
system and we will not discuss them further.

\paragraph{Open and Closed Regions}
A transferable region must be explicitly \emph{opened} using \name's \C{open} construct
in order to either read or update or allocate objects in the region.
Specifically, the construct ``\C{open rgn as v@R S}'' does the following:
(a). It opens the transferable region handled by \C{rgn} for allocation
(i.e., makes it the current allocation context),
(b). binds the identifier \C{R} to this open region, and
(c). initializes the newly introduced local variable \C{v} to refer
to the root object of the region.
The \C{@R} part of the statement is optional and may be omitted.
The \C{open} construct is intended to simplify the problem of ensuring memory safety,
as will be explained soon.
We refer to a transferable region that has not been opened as a \emph{closed} region.

\paragraph{Motivating Example}
% We now illustrate how the features of \name can be used to 
Fig.~\ref{fig:motivating-eg-in-broom} shows how the motivating example
of Fig.~\ref{fig:motivating-eg} can be written in \name.
The \C{onReceive} method receives its input message in a \emph{transferred}
region (\ie, a \emph{closed} region whose ownership is transferred to the
recipient).
Line 7 creates a new region to store the output for time \C{t},
initializing it to contain an empty list.
Line 9 opens the input region to process it.
Line 10 creates a stack region \C{R0}.
Thus, the temporary objects created by the iteration in line 11,
for example, will be allocated in this stack region that lives just
long enough.
We open the desired output region in line 12, so that the new output
objects created by the invocation of \C{selector} in line 13
are allocated in the output region.
Finally, the input region is freed in line 19.

@@

Enforcing this restriction is simple in the case of stack regions since the outlives relation
between stack regions can be inferred from their lexical nesting. Unfortunately,
inferring outlives relations between transferable regions is not easy.
\name imposes the following protocol on the use of transferable regions to help simplify
this check.

@@

\begin{figure}
\includegraphics[scale=0.45]{region-fsm.png}
\caption{The lifetime of a dynamic (transferable) region in \name}
\label{fig:region-fsm}
\end{figure}

The protocol for transferable regions is presented as a finite state machine in Fig.~\ref{fig:region-fsm}.
A transferable region starts its lifetime in a \emph{closed} state, when it is created.
%
The safety of memory accesses in \name is now subject to the condition that
every transferable region correctly follows the state transition discipline shown
in Fig.~\ref{fig:region-fsm}. If this is guaranteed, then \name's region type system
statically guarantees the safety of all memory accesses.
% In other words, the type system reduces
% the problem of ensuring memory safety in \name programs to the problem
% of enforcing the state transition discipline for transferable regions.

In \name, this enforcement is done at runtime by explicitly keeping
track of the \emph{current state} for $\RgnZ$ objects, and
checking the validity of every open, transfer, or free operation
and throwing an exception if it is invalid.
The challenge in enforcing this discipline statically is that transferable regions
are first-class objects. Hence, the program can create multiple aliases for
the same region, \eg, open it via one alias and free it via another.
Typestate verification in the presence of aliases is hard.
The checking can be done statically by preventing the creation of aliases using, \eg, linear types
or unique types. However, this would be quite restrictive, in terms of expressiveness.

\name chooses a reasonable tradeoff. Regions are coarse-grained objects, manipulated relatively infrequently,
in comparison to manipulations of the fine-grained objects that reside inside regions. Hence, the programmer burden
as well as the runtime overhead of checking the region's state transition discipline is acceptable.

\paragraph{Cloning}
Note that the example in Fig.~\ref{fig:motivating-eg-in-broom} works correctly only if \C{selector}
creates an output object that does not share any subobjects with its input object (which resides in
the input region and will be freed at the end of the method).
If there is a need for the output object to point to subobjects of the input object, such subobjects
must be cloned (to copy them from the input region to the output region).
Fortunately, \name's region type system (\S~\ref{sec:type-system}) is
capable of capturing such nuances in the type of \C{selector}
and the type checker will ensure correctness.
Furthermore, the type can be automatically inferred by \name's region type
inference (\S~\ref{sec:type-inference}), which can peform the above
non-trivial reasoning on behalf of the programmer.

%\begin{codejava}
%  Region<string> rgn = new Region<List<String>>
%                        (() => new List<String>());
%  open rgn as  strList@R0 {
%    strList.addAtHead("World");
%    strList.addAtHead("Hello");
%  }
%  rgn.transfer();
%\end{codejava}

>>>>

\paragraph{Stack Regions} The ``\C{letregion R \{ S \}}'' construct
creates a new stack region, with a static identifier \C{R}, whose
scope is restricted to the statement \C{S}. The semantics of
\C{letregion} is similar to Tofte and Talpin~\cite{tofte94}'s
\C{letregion} expression: objects can be allocated by \C{S} in the
newly created region while \C{R} is in scope, but the region and all
objects allocated within it are freed at the end of \C{S}.

@@

The \C{Region} constructor takes as a parameter a function that
constructs the root object: it creates a new region and invokes this
function, with the new region as the allocation context, to create the
root object of the region. The following code illustrates the
creation of a transferable region, whose root is an object of type
\C{A}.
% (\C{r}) of type \C{T}:
\begin{codejava} 
  Region<A> rgn = new Region<T>(() => new A())
\end{codejava} 
In the above code, \C{rgn} is called the \emph{handler} to the newly
created region, and is required to read the contents of the region, or
change its state. The class \C{Region} offers two methods: a \C{free}
method that deallocates the region (and all the objects allocated
within it), and \C{transfer} method that transfers the region to a
consumer process. It is an abstraction of two possible forms of
transfer: a transfer between two processes in a shared memory setting
or a transfer between two processes in a distributed, message-passing,
setting. The precise semantics of \C{transfer} are unimportant in the
context of the region type system and we will not discuss them
further.

% \begin{figure} \begin{codejava} public class Region<T> { public
% Region(Func<T> mkRoot); public void free(); public void transfer();
% } \end{codejava} \caption{The type signature of the \C{Region}
% class} \label{fig:region-class} \end{figure}

% The type signature of the class \C{Region} appears in
% Fig.~\ref{fig:region-class}.  The method \C{free} deallocates the
% region (including all objects allocated within it).  The method
% \C{transfer} transfers the region to a downstream actor as
% determined by the run-time.  It is an abstraction of two possible
% forms of  transfer: a transfer between two actors in a shared memory
% setting or a transfer between two actors in a distributed,
% message-passing, setting.  The precise semantics of \C{transfer} are
% unimportant in the context of the region type system and we will not
% discuss them further.

\paragraph{Open and Closed Regions} A transferable region must be
explicitly \emph{opened} using \name's \C{open} construct in order to
either read or update or allocate objects in the region.
Specifically, the construct ``\C{open rgn as v@R \{ S \}}'' does the
following: (a). It opens the transferable region handled by \C{rgn}
for allocation (i.e., makes it the current allocation context), (b).
binds the identifier \C{R} to this open region, and (c). initializes
the newly introduced local variable \C{v} to refer to the root object
of the region.  The \C{@R} part of the statement is optional and may
be omitted.  The \C{open} construct is intended to simplify the
problem of ensuring memory safety, as will be explained soon.  We
refer to a transferable region that has not been opened as a
\emph{closed} region.

\paragraph{Motivating Example}
% We now illustrate how the features of \name can be used to 
Fig.~\ref{fig:motivating-eg-in-broom} shows how the motivating example
of Fig.~\ref{fig:motivating-eg} can be written in \name.  The
\C{onReceive} method receives its input message in a
\emph{transferred} region (\ie, a \emph{closed} region whose ownership
is transferred to the recipient).  Line 7 creates a new region to
store the output for time \C{t}, initializing it to contain an empty
list.  Line 9 opens the input region to process it.  Line 10 creates a
stack region \C{R0}.  Thus, the temporary objects created by the
iteration in line 11, for example, will be allocated in this stack
region that lives just long enough.  We open the desired output region
in line 12, so that the new output objects created by the invocation
of \C{selector} in line 13 are allocated in the output region.
Finally, the input region is freed in line 15. The output region at
\C{map[t]} stays as along as input messages with timestamp \C{t} keep
arriving. When the timing message for \C{t} arrives, the \C{onNotify}
method transfers the \C{outRgn} at \C{map[t]} to a downstream actor.
After the transfer, \C{SelectVertex} is no longer allowed to access
the \C{outRgn}, facilitating its deallocation.

@@

Enforcing this restriction is simple in the case of stack regions
since the outlives relation between stack regions can be inferred from
their lexical nesting. Unfortunately, inferring outlives relations
in presence of transferable regions is not easy.  \name imposes the following
protocol on the use of transferable regions to help simplify this
check.

@@

\begin{figure} 
\includegraphics[scale=0.45]{region-fsm.png}
\caption{The lifetime of a dynamic (transferable) region in \name}
\label{fig:region-fsm} 
\vspace*{-0.25in} 
\end{figure}

The protocol for transferable regions is presented as a finite state
machine in Fig.~\ref{fig:region-fsm}.  
% A transferable region starts
% its lifetime in a \emph{closed} state, when it is created.
%
The safety of memory accesses in \name is now subject to the condition
that every transferable region correctly follows the state transition
discipline of Fig.~\ref{fig:region-fsm}. Under this condition, \name's
region type system statically guarantees the safety of all memory
accesses.
% In other words, the type system reduces the problem of ensuring
% memory safety in \name programs to the problem of enforcing the
% state transition discipline for transferable regions.

In \name, this enforcement is done at runtime by explicitly keeping
track of the \emph{current state} for $\RgnZ$ objects, and checking
the validity of every open, transfer, or free operation and throwing
an exception if it is invalid.  
% The challenge in enforcing this
% discipline statically is that transferable regions are first-class
% objects. Hence, the program can create multiple aliases for the same
% region, \eg, open it via one alias and free it via another.  Typestate
% verification in the presence of aliases is hard.  The checking can be
% done statically by preventing the creation of aliases using, \eg,
% linear types or unique types. However, this would be quite
% restrictive, in terms of expressiveness.
As explained previously, this is a reasonable tradeoff in the context
of \name, as regions are coarse-grained objects, which are manipulated
infrequently, when compared to fine-grained objects that reside inside
these regions. Therefore, runtime overhead of checking the region's
state transition discipline is acceptable.

\paragraph{Cloning} Note that in the example from
Fig.~\ref{fig:motivating-eg-in-broom} the object returned by the
\C{selector} (on Line 13) should not contain any references to the
input object, since the input region, where the object resides, will
be freed at the end of the method. If there is a need for the output
object to point to subobjects of the input object, such subobjects
must be cloned (to copy them from the input region to the output
region).  Fortunately, \name's region type system
(\S~\ref{sec:type-system}) is capable of capturing such nuances in the
type of \C{selector} and the type checker will ensure correctness.
Furthermore, the type can be automatically inferred by \name's region
type inference (\S~\ref{sec:type-inference}), which can peform the
above reasoning on behalf of the programmer.

%\begin{codejava} Region<string> rgn = new Region<List<String>> (() =>
%new List<String>()); open rgn as  strList@R0 {
%strList.addAtHead("World"); strList.addAtHead("Hello"); }
%rgn.transfer(); \end{codejava}
